---
title: "Forecasting: principles and practice"
author: "Rob J Hyndman"
date: "1.1&nbsp; Time series graphics"
fontsize: 14pt
output:
  beamer_presentation:
    theme: Monash
    keep_tex: no
    toc: false
    fig_width: 7
    fig_height: 3.5
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, warning=FALSE, message=FALSE)
library(fpp2)
source("nicefigs.R")
```


# Time series in R

## `ts` objects and `ts` function

\small

A time series is stored in a `ts` object in R:

 - a list of numbers
 - information about times those numbers were recorded.

###Example

```{r tstable, echo=FALSE, cache=TRUE}
x <- c(123,39,78,52,110)
yr <- 2012:2016
knitr::kable(data.frame(Year=yr,Observation=x), booktabs=TRUE)
```

```r
y <- ts(c(123,39,78,52,110), start=2012)
```

## `ts` objects and `ts` function

For observations that are more frequent than once per year, add a `frequency` argument.

E.g., monthly data stored as a numerical vector `z`:

```r
y <- ts(z, frequency=12, start=c(2003, 1))
```

## `ts` objects and `ts` function

### `ts(data, frequency, start)`

\begin{tabular}{lrl}
\bf Type of data & \hspace*{1.95cm}\bf frequency                 & \bf start example\hspace*{0.25cm} \\
\midrule
Annual           & \only<2->{1}                                  & \only<3->{1995}\\
Quarterly        & \only<4->{4}                                  & \only<5->{c(1995,2)}\\
Monthly          & \only<6->{12}                                 & \only<7->{c(1995,9)}\\
Daily            & \only<8->{7 \emph{or} 365.25}                 & \only<9->{1 \emph{or} c(1995,234)} \\
Weekly           & \only<10->{52.18}                             & \only<11->{c(1995,23)}\\
Hourly           & \only<12->{24 \emph{or} 168 \emph{or} 8,766}  & \only<13->{1}\\
Half-hourly      & \only<14->{48 \emph{or} 336 \emph{or} 17,532} & \only<15>{1}
\end{tabular}

## Australian GDP

```
ausgdp <- ts(x, frequency=4, start=c(1971,3))
```

\vspace*{-0.4cm}

  * Class: "ts"
  * Print and plotting methods available.


```{r, echo=TRUE}
ausgdp
```

## Australian GDP

```{r, echo=TRUE, fig.height=4}
autoplot(ausgdp)
```

## Residential electricity sales


```{r}
elecsales
```



# Time plots

## Time plots

\small

```{r, echo=TRUE, fig.height=4}
autoplot(melsyd[,"Economy.Class"])
```

## Time plots

\small

```{r a10, echo=TRUE}
autoplot(a10) + ylab("$ million") + xlab("Year") +
         ggtitle("Antidiabetic drug sales")
```


#Lab session 1

##

\fontsize{48}{60}\sf\centering
\textbf{Lab Session 1}



# Seasonal plots

## Seasonal plots

\small

```{r, echo=TRUE, fig.height=3.5}
ggseasonplot(a10, ylab="$ million",
  year.labels=TRUE, year.labels.left=TRUE) +
  ggtitle("Seasonal plot: antidiabetic drug sales")
```


## Seasonal plots


  * Data plotted against the individual "seasons" in which the data were observed.  (In this case a "season" is a month.)
  * Something like a time plot except that the data from each season are overlapped.
  * Enables the underlying seasonal pattern to be seen more clearly, and also allows any substantial departures from the seasonal pattern to be easily identified.
  * In R: `ggseasonplot`

## Seasonal polar plots

\small

```{r, fig.height=3.}
ggseasonplot(a10, polar=TRUE) + ylab("$ million")
```

## Seasonal subseries plots

\small

```{r, echo=TRUE, fig.height=3.5}
ggsubseriesplot(a10) + ylab("$ million") +
  ggtitle("Seasonal subseries plot: antidiabetic drug sales")
```



## Seasonal subseries plots

  * Data for each season collected together in time plot as separate time series.
  * Enables the underlying seasonal pattern to be seen clearly, and changes in seasonality over time to be visualized.
  * In R: `ggsubseriesplot`




## Quarterly Australian Beer Production

```{r}
beer <- window(ausbeer,start=1992)
autoplot(beer)
```


## Quarterly Australian Beer Production

```{r}
ggseasonplot(beer,year.labels=TRUE)
```


## Quarterly Australian Beer Production

```{r}
ggsubseriesplot(beer)
```


# Seasonal or cyclic?

##Time series patterns

Trend
  : pattern exists when there is a long-term increase or decrease in the data.
  
Seasonal
  : pattern exists when a series is influenced by seasonal factors (e.g., the quarter of the year, the month, or day of the week).
  
Cyclic
  : pattern exists when data exhibit rises and falls that are \emph{not of fixed period} (duration usually of at least 2 years).


##Time series components

###Differences between seasonal and cyclic patterns:


* seasonal pattern constant length; cyclic pattern variable length
* average length of cycle longer than length of seasonal pattern
* magnitude of cycle more variable than magnitude of seasonal pattern



## Time series patterns

```{r}
autoplot(window(elec, start=1980)) +
  ggtitle("Australian electricity production") +
  xlab("Year") + ylab("GWh")
```

## Time series patterns

```{r}
autoplot(bricksq) +
  ggtitle("Australian clay brick production") +
  xlab("Year") + ylab("million units")
```


## Time series patterns

```{r}
autoplot(hsales) +
  ggtitle("Sales of new one-family houses, USA") +
  xlab("Year") + ylab("Total sales")
```

## Time series patterns

```{r}
autoplot(ustreas) +
  ggtitle("US Treasury Bill Contracts") +
  xlab("Day") + ylab("price")
```

## Time series patterns

```{r}
autoplot(lynx) +
  ggtitle("Annual Canadian Lynx Trappings") +
  xlab("Year") + ylab("Number trapped")
```

## Seasonal or cyclic?

\alert{Differences between seasonal and cyclic patterns:}

  * seasonal pattern constant length; cyclic pattern variable length
  * average length of cycle longer than length of seasonal pattern
  * magnitude of cycle more variable than magnitude of seasonal pattern

\pause

\begin{alertblock}{}
The timing of peaks and troughs is predictable with seasonal data, but unpredictable in the long term with cyclic data.
\end{alertblock}


# Lag plots and autocorrelation

## Example: Beer production

```r
beer <- window(ausbeer, start=1992)
gglagplot(beer, lags=9, do.lines=FALSE, 
  continuous=FALSE)
```

## Example: Beer production

```{r, echo=FALSE, fig.height=6, fig.width=6}
beer <- window(ausbeer, start=1992)
gglagplot(beer, lags=9, do.lines=FALSE, continuous=FALSE)
```

## Lagged scatterplots

  * Each graph shows $y_t$ plotted against $y_{t-k}$ for
different values of $k$.
  * The autocorrelations are the correlations associated
with these scatterplots.

## Autocorrelation

**Covariance** and **correlation**: measure extent of **linear relationship** between two variables ($y$ and $X$).\pause

**Autocovariance** and **autocorrelation**: measure linear relationship between **lagged values** of a time series $y$.\pause

\begin{tabbing}
We measure the relationship between:~~\=  $y_t$ and $y_{t-1}$\\
\> $y_t$ and $y_{t-2}$\\
\> $y_t$ and $y_{t-3}$\\
\> etc.
\end{tabbing}




## Autocorrelation

We denote the sample autocovariance at lag $k$ by $c_k$ and the sample autocorrelation at lag $k$ by $r_k$.  Then define

\begin{block}{}\vspace*{-0.6cm}
\begin{align*}
c_k &= \frac{1}{T}\sum_{t=k+1}^T (y_t-\bar{y})(y_{t-k}-\bar{y}) \\[0.cm]
\text{and}\qquad
r_{k} &= c_k/c_0
\end{align*}\vspace*{-0.8cm}
\end{block}\pause

  * $r_1$ indicates how successive values of  $y$  relate to each other
  * $r_2$ indicates how  $y$ values two periods apart relate to each other
  * $r_k$ is \textit{almost} the same as the sample correlation between $y_t$ and $y_{t-k}$.


## Autocorrelation

\small
Results for first 9 lags for beer data:\footnotesize

```{r, echo=FALSE}
beeracf <- matrix(acf(c(beer), lag.max=9,
                      plot=FALSE)$acf[-1,,1], nrow=1)
colnames(beeracf) <- paste("$r_",1:9,"$",sep="")
knitr::kable(beeracf, booktabs=TRUE,
             align="c", digits=3,
             format.args=list(nsmall=3))
```

```{r beeracf, fig.height=2.7}
ggAcf(beer)
```


## Autocorrelation

  * $r_{4}$  higher than for the other lags. This is due to **the seasonal pattern in the data**: the peaks tend to be **4 quarters** apart and the troughs tend to be **2 quarters** apart.
  * $r_2$ is more negative than for the other lags because troughs tend to be 2 quarters behind peaks.
  * Together, the autocorrelations at lags 1, 2, \dots, make up the \emph{autocorrelation} or ACF.
  * The plot is known as a **correlogram**



## Recognizing seasonality in a time series

If there is seasonality, the ACF at the seasonal lag (e.g., 12 for monthly data) will be **large and positive**.

  * For seasonal monthly data, a large ACF value will be seen at lag 12 and possibly also at lags 24, 36, \dots
  * For seasonal quarterly data, a large ACF value will be seen at lag 4 and possibly also at lags 8, 12, \dots

## Aus monthly electricity production

```{r}
elec2 <- window(elec, start=1980)
autoplot(elec2)
```

## Aus monthly electricity production

```{r}
ggAcf(elec2, lag.max=48)
```

## Aus monthly electricity production

Time plot shows clear trend and seasonality.

The same features are reflected in the ACF.

  * The slowly decaying ACF indicates trend.
  * The ACF peaks at lags 12, 24, 36, \dots, indicate seasonality of length 12.

## Which is which?

```{r, fig.height=6, fig.width=12, echo=FALSE}
tp1 <- autoplot(cowtemp) + xlab("") + ylab("chirps per minute") +
  ggtitle("1. Daily temperature of cow")
tp2 <- autoplot(USAccDeaths/1e3) + xlab("") + ylab("thousands") +
  ggtitle("2. Monthly accidental deaths")
tp3 <- autoplot(AirPassengers) + xlab("") + ylab("thousands") +
  ggtitle("3. Monthly air passengers")
tp4 <- autoplot(mink/1e3) + xlab("") + ylab("thousands") +
  ggtitle("4. Annual mink trappings")
acfb <- ggAcf(cowtemp, ci=0) + xlab("") + ggtitle("B") + ylim(-0.4,1)
acfa <- ggAcf(USAccDeaths, ci=0) + xlab("") + ggtitle("A") + ylim(-0.4,1)
acfd <- ggAcf(AirPassengers, ci=0) + xlab("") + ggtitle("D") + ylim(-0.4,1)
acfc <- ggAcf(mink, ci=0) + xlab("") + ggtitle("C") + ylim(-0.4,1)
gridExtra::grid.arrange(tp1,tp2,tp3,tp4,
                        acfa,acfb,acfc,acfd,nrow=2)
```

# White noise

## Example: White noise

```{r, fig.height=3.5}
wn <- ts(rnorm(36))
autoplot(wn)
```

\only<2>{
\begin{textblock}{10}(1.4,6)
\begin{block}{}
White noise data is uncorrelated across time with zero mean and constant variance.

(Technically, we require independence as well.)
\end{block}
\end{textblock}}

## Example: White noise

```{r, results='asis', echo=FALSE}
wnacf <- matrix(acf(c(wn), lag.max=10,
                      plot=FALSE)$acf[-1,,1], nrow=1)
colnames(wnacf) <- paste("$r_{",1:10,"}$",sep="")
print(xtable::xtable(t(wnacf)),
    sanitize.rownames.function=identity,
    booktabs=TRUE,
    include.colnames = FALSE,
    hline.after = FALSE,
    size='small',
    comment=FALSE,
    floating=FALSE)
```

```{r, echo=FALSE}
savepdf("wnacf")
ggAcf(wn, max.lag=15)
endpdf()
```

\placefig{4.5}{1.6}{width=8cm}{wnacf}

\centerline\textbf{Sample autocorrelations for white noise series.}

For uncorrelated data, we would expect each
autocorrelation to be close to zero.



## \large Sampling distribution of autocorrelations

Sampling distribution of $r_k$ for white noise data is asymptotically N(0,$1/T$).\pause

  *  95% of all $r_k$ for white noise must lie within $\pm 1.96/\sqrt{T}$.
  * If this is not the case, the series is probably not WN.
  * Common to plot lines at $\pm 1.96/\sqrt{T}$ when plotting ACF.
These are the \textcolor{orange}{\textbf{\emph{critical values}}}.



## Autocorrelation

\placefig{5}{1.6}{width=8cm}{wnacf}

\begin{textblock}{4.8}(0.2,1.5)
\structure{Example:}

$T=36$ and so critical values at $\pm
1.96/\sqrt{36} = \pm 0.327$.

All autocorrelation coefficients lie within these
limits, confirming  that the data are white noise. (More precisely, the data cannot be \\
distinguished \rlap{from white noise.)}
\end{textblock}




## Example: Pigs slaughtered
\fontsize{13}{13}\sf

```{r}
pigs2 <- window(pigs, start=1990)
autoplot(pigs2) +
  xlab("Year") + ylab("thousands") +
  ggtitle("Number of pigs slaughtered in Victoria")
```

## Example: Pigs slaughtered

```{r}
ggAcf(pigs2)
```

## Example: Pigs slaughtered

Monthly total number of pigs slaughtered
in the state of Victoria, Australia, from January 1990 through August 1995.
(Source: Australian Bureau of Statistics.)\pause


  * Difficult to detect pattern in time plot.
  * ACF shows some significant autocorrelation at lags 1, 2, and 3.
  * $r_{12}$ relatively large although not significant.  This may indicate
some slight seasonality.

\pause

These show the series is **not a white noise series**.

## Combination graph

```{r, echo=TRUE, fig.height=3.6}
ggtsdisplay(pigs2, plot.type='scatter')
```

#Lab session 2

##

\fontsize{48}{60}\sf\centering
\textbf{Lab Session 2}

