---
title: "Forecasting: principles and practice"
author: "Rob J Hyndman"
date: "1.3&nbsp; Seasonality and trends"
fontsize: 14pt
output:
  beamer_presentation:
    fig_width: 7
    fig_height: 3.5
    highlight: tango
    theme: metropolis
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache=TRUE, warning=FALSE, message=FALSE)
library(fpp2)
```


# Time series components

##Time series patterns

Trend
  : pattern exists when there is a long-term increase or decrease in the data.

Seasonal
  : pattern exists when a series is influenced by seasonal factors (e.g., the quarter of the year, the month, or day of the week).

Cyclic
  : pattern exists when data exhibit rises and falls that are \emph{not of fixed period} (duration usually of at least 2 years).


##Time series decomposition

\begin{block}{}\vspace*{-0.3cm}
\[ Y_t = f(S_t, T_t, E_t) \]
\end{block}
\begin{tabular}{@{}llp{8cm}@{}}
where & $Y_t=$ & data at period $t$ \\
      & $S_t=$ & seasonal component at period $t$ \\
      & $T_t=$ & trend-cycle component at period $t$\\
      & $E_t=$ & remainder (or irregular or error) component at period $t$
\end{tabular}
\pause
\vspace*{0.5cm}

**Additive decomposition**:
$Y_t = S_t + T_t + E_t.$
\pause

**Multiplicative decomposition:**
$Y_t = S_t \times T_t \times E_t.$



##Time series decomposition

* Additive model appropriate if magnitude of seasonal fluctuations does not vary with level.

* If seasonal are proportional to level of series, then multiplicative model appropriate.
* Multiplicative decomposition more prevalent with economic series

* Alternative: use a Box-Cox transformation, and then use additive decomposition.

* Logs turn multiplicative relationship into
an additive relationship:
\[\hspace*{-0.7cm}
Y_t = S_t \times T_t \times E_t \quad\Rightarrow\quad
\log Y_t = \log S_t + \log T_t + \log E_t.
\]



##Euro electrical equipment

```{r}
fit <- stl(elecequip,s.window=5)
tab <- cbind(Data=elecequip, Trend=fit$time.series[,2])
autoplot(tab,main="Electrical equipment manufacturing (Euro area)",ylab="New orders index",xlab="")
```

##Euro electrical equipment

```{r}
autoplot(fit)
```

##Euro electrical equipment

```{r}
ggmonthplot(fit$time.series[,"seasonal"], main="", ylab="Seasonal")
```


##Seasonal adjustment


* Useful by-product of decomposition:  an easy way to
calculate seasonally adjusted data.

* Additive decomposition: seasonally adjusted data given by
$$Y_t - S_t = T_t + E_t$$

* Multiplicative decomposition: seasonally adjusted data given by
$$Y_t / S_t = T_t \times E_t$$



##Euro electrical equipment

```{r}
tab <- cbind(Data=elecequip, SeasAdjust=seasadj(fit))
autoplot(tab,main="Electrical equipment manufacturing",xlab="",ylab="New orders index")
```


##Seasonal adjustment

* We use estimates of $S$ based on past values to seasonally  adjust
a current value.

* Seasonally adjusted  series reflect \textbf{remainders} as well as \textbf{trend}. Therefore they are not "smooth" and "downturns" or "upturns" can be misleading.

* It is better to use the trend-cycle component to look for turning
points.





##\normalsize History of time series decomposition


* Classical method originated in 1920s.

* Census II method introduced in 1957. Basis for modern X-12-ARIMA method.

* STL method introduced in 1983

* TRAMO/SEATS introduced in 1990s.







# STL decomposition

##STL decomposition

* STL: "Seasonal and Trend decomposition using Loess",
* Very versatile and robust.
* Unlike X-12-ARIMA, STL will handle any type of seasonality.
* Seasonal component allowed to change over time, and rate of change controlled by user.
* Smoothness of trend-cycle also controlled by user.
* Robust to outliers
* Not trading day or calendar adjustments.
* Only additive.


##Euro electrical equipment
\fontsize{12}{14}\sf

```{r, echo=TRUE, fig.height=3.6}
elecequip %>% stl(s.window=5) %>%
  autoplot
```

##Euro electrical equipment
\fontsize{12}{14}\sf

```{r, echo=TRUE, fig.height=3.6}
elecequip %>%
  stl(t.window=15, s.window='periodic', robust=TRUE) %>%
  autoplot
```

##STL decomposition in R

* `t.window` controls wiggliness of trend component.
* `s.window` controls variation on seasonal component.


# Forecasting and decomposition

##Forecasting and decomposition


* Forecast seasonal component by repeating the last year
* Forecast seasonally adjusted data using non-seasonal time series method. E.g.,
    - Holt's method --- next topic
    - Random walk with drift model
* Combine forecasts of seasonal component with forecasts of seasonally adjusted data to get forecasts of original data.
* Sometimes a decomposition is useful just for understanding the data before building a separate forecasting model.

##Seas adj elec equipment

```{r}
fit <- stl(elecequip, t.window=15,
  s.window="periodic", robust=TRUE)
eeadj <- seasadj(fit)
autoplot(naive(eeadj, h=24),
  main="Naive forecasts of seasonally adjusted data",xlab="New orders index")
```

##Seas adj elec equipment

```{r}
fcast <- forecast(fit,method="naive", h=24)
autoplot(fcast,ylab="New orders index")

```


##How to do this in R

```r
fit <- stl(elecequip, t.window=15,
  s.window="periodic", robust=TRUE)

eeadj <- seasadj(fit)
autoplot(naive(eeadj, h=24)) +
  ylab("New orders index")

fcast <- forecast(fit, method="naive", h=24)
autoplot(fcast) +
  ylab="New orders index")
```


##\normalsize Decomposition and prediction intervals


* It is common to take the prediction intervals from the seasonally adjusted forecasts and modify them with the seasonal component.

* This ignores the uncertainty in the seasonal component estimate.

* It also ignores the uncertainty in the future seasonal pattern.



##Some more R functions

```r
fcast <- stlf(elecequip, method='naive')

fcast <- stlf(elecequip, method='naive',
  h=36, s.window=11, robust=TRUE)

```

#Lab session 5
##

\fontsize{48}{60}\sf\centering
\textbf{Lab Session 5 }

