---
title: "Forecasting: principles and practice"
author: "Rob J Hyndman"
date: "2.5&nbsp; Seasonal ARIMA models"
fontsize: 14pt
output:
  beamer_presentation:
    theme: Monash
    keep_tex: no
    toc: false
    fig_width: 7
    fig_height: 4.3
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache=TRUE, warning=FALSE, message=FALSE)
library(fpp2)
```

#Backshift notation reviewed

##Backshift notation

A  very useful notational device is the backward  shift operator,  $B$,  which is used as follows:
$$
{B y_{t}  =  y_{t - 1}} \: .
$$\pause
In  other words, $B$, operating on $y_{t}$, has the effect of **shifting the  data back one period**. \pause
Two applications of  $B$  to  $y_{t}$ **shifts the data  back  two periods**:
$$
B(By_{t})  =  B^{2}y_{t}  =  y_{t-2}\: .
$$\pause
For  monthly  data, if we wish to shift attention  to  ``the same  month last year,''  then  $B^{12}$ is used,  and  the notation is  $B^{12}y_{t}$  =  $y_{t-12}$.


##Backshift notation

  * First difference: $1-B$.
  * Double difference:  $(1- B)^{2}$.
  * $d$th-order difference: $(1 - B)^{d} y_{t}.$
  * Seasonal difference: $1-B^m$.
  * Seasonal difference followed by a first difference: $(1-B)(1-B^m)$.
  * Multiply terms together together to see the combined effect:
\begin{align*}
(1-B)(1-B^m)y_t &= (1 - B - B^m + B^{m+1})y_t \\
&= y_t-y_{t-1}-y_{t-m}+y_{t-m-1}.
\end{align*}

##Backshift notation for ARIMA
\fontsize{13}{15}\sf

  * ARMA model:

  \vspace*{-0.9cm}
\begin{align*}
\hspace*{0.5cm} y_{t}  &=  c  +  \phi_{1}y_{t - 1}  +  \cdots  +  \phi_{p}y_{t - p}
 + e_t + \theta_{1}e_{t - 1} +  \cdots  + \theta_{q}e_{t - q}\\
       &=  c + \phi_{1}By_{t} + \cdots + \phi_pB^py_{t}
           +  e_{t}  +  \theta_{1}Be_{t} + \cdots + \theta_qB^qe_{t} \\
\phi(B)y_t & = c + \theta(B) e_t\\
	& \hspace*{0.6cm} \text{where $\phi(B)= 1-\phi_1B - \cdots - \phi_p B^p$}\\
	& \hspace*{0.6cm} \text{and $\theta(B) = 1 + \theta_1 B + \cdots + \theta_q B^q$.}
\end{align*}

\pause

  * ARIMA(1,1,1) model:

$$
\begin{array}{c c c c}
(1 - \phi_{1} B) & (1  -  B) y_{t} &= &c + (1  + \theta_{1} B) e_{t}\\
\uparrow  & \uparrow    &   &\uparrow\\
{\text{AR(1)}} & {\text{First}}   &     &{\text{MA(1)}}\\
& {\hbox to 0cm{\hss\text{difference}\hss}}\\
\end{array}
$$


##Backshift notation for ARIMA
\fontsize{13}{15}\sf


  * ARIMA($p,d,q$) model:


\begin{equation*}
  \arraycolsep=0.1cm
  \begin{array}{c c c c}
    (1-\phi_1B - \cdots - \phi_p B^p) & (1-B)^d y_{t} &= &c + (1 + \theta_1 B + \cdots + \theta_q B^q)e_t\\
    {\uparrow} & {\uparrow} & &{\uparrow}\\
    {\text{AR($p$)}} & \hbox to 0cm{\hss\text{$d$ differences}\hss} & &{\text{MA($q$)}}\\
  \end{array}
\end{equation*}



#Seasonal ARIMA models

##Seasonal ARIMA models

| ARIMA | $~\underbrace{(p, d, q)}$ | $\underbrace{(P, D, Q)_{m}}$ |
| ----: | :-----------------------: | :--------------------------: |
|       | ${\uparrow}$              | ${\uparrow}$                 |
|       | Non-seasonal part         | Seasonal part of             |
|       | of the model              | of the model                 |


where $m =$ number of observations per year.



##Seasonal ARIMA models

E.g., ARIMA$(1, 1, 1)(1, 1, 1)_{4}$  model (without constant)\pause
$$(1 - \phi_{1}B)(1 - \Phi_{1}B^{4}) (1 - B) (1 - B^{4})y_{t} ~= ~
(1 + \theta_{1}B) (1 + \Theta_{1}B^{4})e_{t}.
$$\pause

\setlength{\unitlength}{1mm}
\begin{footnotesize}
\begin{picture}(100,25)(-5,0)
\thinlines
{\put(5,22){\vector(0,1){6}}}
{\put(22,10){\vector(0,1){18}}}
{\put(42,22){\vector(0,1){6}}}
{\put(57,10){\vector(0,1){18}}}
{\put(83,22){\vector(0,1){6}}}
{\put(105,10){\vector(0,1){18}}}
{\put(-10,17){$\left(\begin{array}{@{}c@{}} \text{Non-seasonal} \\ \text{AR(1)}
                    \end{array}\right)$}}
{\put(12,5){$\left(\begin{array}{@{}c@{}} \text{Seasonal} \\ \text{AR(1)}
                    \end{array}\right)$}}
{\put(25,17){$\left(\begin{array}{@{}c@{}} \text{Non-seasonal} \\ \text{difference}
                    \end{array}\right)$}}
{\put(45,5){$\left(\begin{array}{@{}c@{}} \text{Seasonal} \\ \text{difference}
                    \end{array}\right)$}}
{\put(70,17){$\left(\begin{array}{@{}c@{}} \text{Non-seasonal} \\ \text{MA(1)}
                    \end{array}\right)$}}
{\put(95,5){$\left(\begin{array}{@{}c@{}} \text{Seasonal} \\ \text{MA(1)}
                    \end{array}\right)$}}
\end{picture}
\end{footnotesize}


\vspace*{10cm}




##Seasonal ARIMA models

E.g., ARIMA$(1, 1, 1)(1, 1, 1)_{4}$  model (without constant)
$$(1 - \phi_{1}B)(1 - \Phi_{1}B^{4}) (1 - B) (1 - B^{4})y_{t} ~= ~
(1 + \theta_{1}B) (1 + \Theta_{1}B^{4})e_{t}.
$$

All the factors can be multiplied out and the general model
written as follows:
\begin{align*}
y_{t}  &= (1 + \phi_{1})y_{t - 1} - \phi_1y_{t-2} + (1 + \Phi_{1})y_{t - 4}\\
&\text{}
 -  (1  + \phi_{1}  +  \Phi_{1} + \phi_{1}\Phi_{1})y_{t - 5}
 +  (\phi_{1}  +  \phi_{1} \Phi_{1}) y_{t - 6} \\
& \text{}  - \Phi_{1} y_{t - 8} +  (\Phi_{1}  +  \phi_{1} \Phi_{1}) y_{t - 9}
  - \phi_{1} \Phi_{1} y_{t  -  10}\\
  &\text{}
+    e_{t} + \theta_{1}e_{t - 1} + \Theta_{1}e_{t - 4}  + \theta_{1}\Theta_{1}e_{t - 5}.
\end{align*}
\vspace*{10cm}




##Common ARIMA models

In the US Census Bureau uses the following models most often:\vspace*{0.5cm}

\begin{tabular}{|ll|}
\hline
ARIMA(0,1,1)(0,1,1)$_m$& with log transformation\\
ARIMA(0,1,2)(0,1,1)$_m$& with log transformation\\
ARIMA(2,1,0)(0,1,1)$_m$& with log transformation\\
ARIMA(0,2,2)(0,1,1)$_m$& with log transformation\\
ARIMA(2,1,2)(0,1,1)$_m$& with no transformation\\
\hline
\end{tabular}




##Seasonal ARIMA models
The seasonal part of an AR or MA model will be seen in the seasonal lags of
the PACF and ACF.

\structure{ARIMA(0,0,0)(0,0,1)$_{12}$ will show:}

  * a spike at lag 12 in the ACF but no other significant spikes.
  * The PACF will show exponential decay in the seasonal lags;
     that is, at lags 12, 24, 36, \dots.

\structure{ARIMA(0,0,0)(1,0,0)$_{12}$ will show:}

  *  exponential decay in the seasonal lags of the ACF
  * a single significant spike at lag 12 in the PACF.

##European quarterly retail trade

```{r, echo=TRUE, fig.height=3.6}
autoplot(euretail) +
  xlab("Year") + ylab("Retail index")
```

##European quarterly retail trade

```{r, echo=TRUE, fig.height=4}
euretail %>% diff(lag=4) %>% ggtsdisplay()
```

##European quarterly retail trade

```{r, echo=TRUE, fig.height=3.8}
euretail %>% diff(lag=4) %>% diff() %>%
  ggtsdisplay()
```

##European quarterly retail trade

  * $d=1$ and $D=1$ seems necessary.
  * Significant spike at lag 1 in ACF suggests non-seasonal MA(1) component.
  * Significant spike at lag 4 in ACF suggests seasonal MA(1) component.
  * Initial candidate model: ARIMA(0,1,1)(0,1,1)$_4$.
  * We could also have started with ARIMA(1,1,0)(1,1,0)$_4$.

##European quarterly retail trade

```{r, echo=TRUE, fig.height=3.5}
fit <- Arima(euretail, order=c(0,1,1),
  seasonal=c(0,1,1))
checkresiduals(fit)
```

##European quarterly retail trade

```{r, echo=FALSE}
checkresiduals(fit, plot=FALSE)
```

##European quarterly retail trade

  * ACF and PACF of residuals show significant spikes at lag 2, and maybe lag 3.
  * AICc of ARIMA(0,1,2)(0,1,1)$_4$ model is 74.36.
  * AICc of ARIMA(0,1,3)(0,1,1)$_4$ model is 68.53.
\pause\vfill

```r
fit <- Arima(euretail, order=c(0,1,3),
  seasonal=c(0,1,1))
checkresiduals(fit)
```


##European quarterly retail trade

\fontsize{12}{15}\sf

```{r}
(fit <- Arima(euretail, order=c(0,1,3),
  seasonal=c(0,1,1)))
```

##European quarterly retail trade
\fontsize{13}{15}\sf

```{r, echo=TRUE, fig.height=4}
checkresiduals(fit)
```

##European quarterly retail trade
\fontsize{13}{15}\sf

```{r, echo=FALSE}
checkresiduals(fit, plot=FALSE)
```

##European quarterly retail trade

```{r, echo=TRUE, fig.height=4}
autoplot(forecast(fit, h=12))
```

##European quarterly retail trade
\fontsize{12}{14}\sf

```{r, echo=TRUE}
auto.arima(euretail)
```

##European quarterly retail trade
\fontsize{12}{14}\sf

```{r euretailtryharder, echo=TRUE}
auto.arima(euretail, stepwise=FALSE, approximation=FALSE)
```


##Cortecosteroid drug sales


```{r h02}
lh02 <- log(h02)
tmp <- cbind("H02 sales (million scripts)" = h02,
             "Log H02 sales"=lh02)
autoplot(tmp, facets=TRUE) + xlab("Year") + ylab("")
```

##Cortecosteroid drug sales
```{r h02b}
ggtsdisplay(diff(lh02,12), xlab="Year",
  main="Seasonally differenced H02 scripts")
```

##Cortecosteroid drug sales

  * Choose $D=1$ and $d=0$.
  * Spikes in PACF at lags 12 and 24 suggest seasonal AR(2) term.
  * Spikes in PACF suggests possible non-seasonal AR(3) term.
  * Initial candidate model: ARIMA(3,0,0)(2,1,0)$_{12}$.


##Cortecosteroid drug sales


| Model | AICc|
|:------: |:------:|
| ARIMA(3,0,0)(2,1,0)$_{12}$ | $-475.12$ |
| ARIMA(3,0,1)(2,1,0)$_{12}$ | $-476.31$ |
| ARIMA(3,0,2)(2,1,0)$_{12}$ | $-474.88$ |
| ARIMA(3,0,1)(1,1,0)$_{12}$ | $-463.40$ |
| ARIMA(3,0,1)(0,1,1)$_{12}$ | $-483.67$ |
| ARIMA(3,0,1)(0,1,2)$_{12}$ | $-485.48$ |
| ARIMA(3,0,1)(1,1,1)$_{12}$ | $-484.25$ |


##Cortecosteroid drug sales
\fontsize{11}{14}\sf

```{r, echo=TRUE}
(fit <- Arima(h02, order=c(3,0,1), seasonal=c(0,1,2),
   lambda=0))
```


##Cortecosteroid drug sales

```{r h02res, echo=TRUE, fig.height=4}
checkresiduals(fit)
```

##Cortecosteroid drug sales

```{r h02resb, echo=FALSE, fig.height=4}
checkresiduals(fit, plot=FALSE)
```

##Cortecosteroid drug sales
\fontsize{10}{14}\sf

```{r h02tryharder, echo=TRUE, fig.height=3.6}
(fit <- auto.arima(h02, lambda=0, d=0, D=1, max.order=9,
  stepwise=FALSE, approximation=FALSE))
```


##Cortecosteroid drug sales
\fontsize{13}{15}\sf

```{r, echo=TRUE, fig.height=4}
checkresiduals(fit)
```


##Cortecosteroid drug sales
\fontsize{13}{15}\sf

```{r, echo=FALSE}
checkresiduals(fit, plot=FALSE)
```

##Cortecosteroid drug sales
\fontsize{11}{12}\sf

Training data: July 1991 to June 2006

Test data: July 2006--June 2008

```r
getrmse <- function(x,h,...)
{
  train.end <- time(x)[length(x)-h]
  test.start <- time(x)[length(x)-h+1]
  train <- window(x,end=train.end)
  test <- window(x,start=test.start)
  fit <- Arima(train,...)
  fc <- forecast(fit,h=h)
  return(accuracy(fc,test)[2,"RMSE"])
}
getrmse(h02,h=24,order=c(3,0,0),seasonal=c(2,1,0),lambda=0)
getrmse(h02,h=24,order=c(3,0,1),seasonal=c(2,1,0),lambda=0)
getrmse(h02,h=24,order=c(3,0,2),seasonal=c(2,1,0),lambda=0)
getrmse(h02,h=24,order=c(3,0,1),seasonal=c(1,1,0),lambda=0)
getrmse(h02,h=24,order=c(3,0,1),seasonal=c(0,1,1),lambda=0)
getrmse(h02,h=24,order=c(3,0,1),seasonal=c(0,1,2),lambda=0)
getrmse(h02,h=24,order=c(3,0,1),seasonal=c(1,1,1),lambda=0)
getrmse(h02,h=24,order=c(4,0,3),seasonal=c(0,1,1),lambda=0)
getrmse(h02,h=24,order=c(3,0,3),seasonal=c(0,1,1),lambda=0)
getrmse(h02,h=24,order=c(4,0,2),seasonal=c(0,1,1),lambda=0)
getrmse(h02,h=24,order=c(3,0,2),seasonal=c(0,1,1),lambda=0)
getrmse(h02,h=24,order=c(2,1,3),seasonal=c(0,1,1),lambda=0)
getrmse(h02,h=24,order=c(2,1,4),seasonal=c(0,1,1),lambda=0)
getrmse(h02,h=24,order=c(2,1,5),seasonal=c(0,1,1),lambda=0)
```

##Cortecosteroid drug sales
\fontsize{12}{14}\sf

```{r, cache=TRUE}
models <- rbind(
  c(3,0,0,2,1,0),
  c(3,0,1,2,1,0),
  c(3,0,2,2,1,0),
  c(3,0,1,1,1,0),
  c(3,0,1,0,1,1),
  c(3,0,1,0,1,2),
  c(3,0,1,1,1,1),
  c(4,0,3,0,1,1),
  c(3,0,3,0,1,1),
  c(4,0,2,0,1,1),
  c(3,0,2,0,1,1),
  c(2,1,3,0,1,1),
  c(2,1,4,0,1,1),
  c(2,1,5,0,1,1))
h <- 24
train.end <- time(h02)[length(h02)-h]
test.start <- time(h02)[length(h02)-h+1]
train <- window(h02,end=train.end)
test <- window(h02,start=test.start)

rmse <- numeric(NROW(models))
modelname <- character(NROW(models))
for(i in seq(length(rmse)))
{
  fit <- Arima(train, order=models[i,1:3],
          seasonal=models[i,4:6], lambda=0)
  fc <- forecast(fit,h=h)
  rmse[i] <- accuracy(fc, test)[2,"RMSE"]
  modelname[i] <- as.character(fit)
}
knitr::kable(data.frame(Model=modelname,RMSE=rmse),
             digits=4)
```


##Cortecosteroid drug sales

  * Models with lowest AICc values tend to give slightly better results than the other models.
  * AICc comparisons must have the same orders of differencing. But RMSE test set comparisons can involve any models.
  * No model passes all the residual tests.
  * Use the best model available, even if it does not pass all tests.
  * In this case, the ARIMA(3,0,1)(0,1,2)$_{12}$ has the lowest RMSE value and the best  AICc value for models with fewer than 6 parameters.



##Cortecosteroid drug sales
\fontsize{12}{14}\sf

```{r h02f, echo=TRUE, fig.height=3}
fit <- Arima(h02, order=c(3,0,1), seasonal=c(0,1,2),
  lambda=0)
autoplot(forecast(fit)) +
  ylab("H02 sales (million scripts)") + xlab("Year")
```



#ARIMA vs ETS


##ARIMA vs ETS


  * Myth that ARIMA models are more general than exponential smoothing.

  * Linear exponential smoothing models all special cases of ARIMA models.

  * Non-linear exponential smoothing models have no equivalent ARIMA counterparts.

  * Many ARIMA models have no exponential smoothing counterparts.

  * ETS models all non-stationary. Models with seasonality or non-damped trend (or both) have two unit roots; all other models have one unit root.





##Equivalences

\fontsize{12}{14}\sf

|**ETS model**  | **ARIMA model**             | **Parameters**                       |
| :------------ | :-------------------------- | :----------------------------------- |
| ETS(A,N,N)    | ARIMA(0,1,1)                | $\theta_1 = \alpha-1$                |
| ETS(A,A,N)    | ARIMA(0,2,2)                | $\theta_1 = \alpha+\beta-2$          |
|               |                             | $\theta_2 = 1-\alpha$                |
| ETS(A,A,N)    | ARIMA(1,1,2)                | $\phi_1=\phi$                        |
|               |                             | $\theta_1 = \alpha+\phi\beta-1-\phi$ |
|               |                             | $\theta_2 = (1-\alpha)\phi$          |
| ETS(A,N,A)    | ARIMA(0,0,$m$)(0,1,0)$_m$   |                                      |
| ETS(A,A,A)    | ARIMA(0,1,$m+1$)(0,1,0)$_m$ |                                      |
| ETS(A,A,A)    | ARIMA(1,0,$m+1$)(0,1,0)$_m$ |                                      |



#Lab session 12
##

\fontsize{48}{60}\sf\centering
\textbf{Lab Session 12}


