---
title: "Lab sessions: Day 3"
author: "Rob J Hyndman"
date: "27 June 2018"
output:
  html_document:
    fig_height: 5
    fig_width: 8
    toc: yes
    toc_depth: 1
    toc_float:
      collapsed: false
    number_sections: false
    theme: readable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, cache=TRUE, messages=FALSE, warnings=FALSE)
library(fpp2)
```

# Lab Session 18

```{r lab18a}
avecost <- motel[,"Takings"] / motel[,"Roomnights"]
```

`cpimel` contains quarterly CPI values. We can use linear approximation to interpolate the quarterly data to obtain monthly CPI.

```{r lab18b}
qcpi <- ts(approx(time(cpimel), cpimel, time(motel), rule=2)$y,
  start=start(motel), frequency=frequency(motel))
```

```{r lab18c}
autoplot(cbind(avecost, qcpi), facets=TRUE)
```

We expect avecost to be related to CPI, but the variance of average cost increases with the level. So logs will help. Also, average cost is likely to be a multiple of CPI as it will depend on lots of individual costs, each of which will increase with CPI. So logarithms will turn the multiplicative relationship into something additive which we can model.

```{r lab18d}
autoplot(log(cbind(avecost, qcpi)), facets=TRUE)
```

```{r lab18e}
(fit <- auto.arima(avecost, xreg=log(qcpi), lambda=0, biasadj=TRUE))
```

We will use an ARIMA model for CPI:

```{r lab18f}
fitcpi <- auto.arima(qcpi)
fccpi <- forecast(fitcpi, h=12)
autoplot(fccpi)
```

Now we can forecast average cost

```{r lab18g}
fc <- forecast(fit, xreg=log(fccpi$mean))
autoplot(fc)
```


# Lab Session 19


```{r retail}
retaildata <- read.csv("retail.csv")
mytimeseries <- ts(retaildata[,4], frequency=12, start=c(1982,4))
train <- window(mytimeseries, end=c(2010,12))
test <- window(mytimeseries, start=2011)
```

```{r lab19a, dependson='retail'}
lambda <- BoxCox.lambda(mytimeseries)
# Find the order of Fourier terms which gives minimum AIC
bestmodel <- list(aicc=Inf)
for(k in seq(6)) {
  fit <- auto.arima(train, lambda=lambda,
    xreg=fourier(train, K=k))
  if(fit$aicc < bestmodel$aicc) {
    bestmodel <- fit
    bestK <- k
  }
}
fc <- forecast(bestmodel, xreg=fourier(train, bestK, 24))
accuracy(fc, test)
```

```{r lab19b, dependson='lab19a'}
checkresiduals(fc)
autoplot(fc)
```

# Lab Session 20

```{r lab20a}
fit <- tbats(mytimeseries)
checkresiduals(fit)
forecast(fit) %>% autoplot
```

```{r lab20b}
gastrain <- subset(gasoline, end=length(gasoline)-2*52)
gastest <- subset(gasoline, start=length(gasoline)-2*52+1)
fit <- tbats(gastrain)
checkresiduals(fit)
fc1 <- forecast(fit)
autoplot(fc1)
accuracy(fc1, gastest)
```

```{r lab20c}
fc2 <- stlf(gastrain)
accuracy(fc2, gastest)
```

```r lab20d}
# Find the order of Fourier terms which gives minimum AIC
bestmodel <- list(aicc=Inf)
for(k in seq(26)) {
  fit <- auto.arima(gastrain,
    xreg=fourier(gastrain, K=k))
  if(fit$aicc < bestmodel$aicc) {
    bestmodel <- fit
    bestK <- k
  }
}
fc3 <- forecast(bestmodel, xreg=fourier(gastrain, bestK, 24))
accuracy(fc3, gastest)
```

# Lab Session 21

```{r lab21a, fig.height=8}
library(hts)
plot(infantgts)
smatrix(infantgts)

# Forecast 10-steps-ahead and reconcile the forecasts
infantforecast <- forecast(infantgts, h=10)

# Plot the forecasts including only the last ten historical years
plot(infantforecast, include=10)

# set up training and testing sets
training <- window(infantgts, end=1993)
test <- window(infantgts, start=1994)

# Compute forecasts on training data
forecast <- forecast(training, h=10)

# calculate ME, RMSE, MAE, MAPE, MPE and MASE
accuracy.gts(forecast, test)
```

```{r lab21b}
# Overall forecast accuracy
rmse <- accuracy(forecast, test)["RMSE",]
sqrt(sum(rmse^2))
```

```{r lab21c}
forecast_bu <- forecast(training, h=10, method="bu")
sqrt(sum(accuracy(forecast_bu, test)["RMSE",]^2))
```


# Lab Session 22


```{r lab22a, dependson='retail'}
library(thief)
train <- window(mytimeseries, end=c(2010,12))
test <- window(mytimeseries, start=2011)
lambda <- BoxCox.lambda(mytimeseries)
fcarima <- forecast(auto.arima(train, lambda=lambda), h=length(test))
fcthief <- thief(train, usemodel='arima', lambda=lambda, h=length(test))
autoplot(mytimeseries) +
  autolayer(fcthief, series="Thief") +
  autolayer(fcarima, series="ARIMA", PI=FALSE) 
```

```{r lab22b, dependson=c('retail','lab22a')}
autoplot(train, include=60) +
  autolayer(fcthief, series="Thief") +
  autolayer(fcarima, series="ARIMA", PI=FALSE) 
```

```{r lab22c, dependson='lab22a'}
accuracy(fcarima, test)
accuracy(fcthief, test)
```

